Challenges to transboundary fisheries management in North America under climate change

### 2021-03-19 Ecol/Manuscript_Draft.Rmd
grammar fix

In 1982, the United Nations Law of the Sea Convention (UNCLOS) formalized the concept of exclusive economic zones (EEZs) creating the concept that we know today as shared stocks [@UnitedNations:1986tl], i.e. stocks that migrate between countries EEZs (known as transboundary stocks) or between EEZs and the high seas (also called straddling stocks) [@Song:2017iua]. Today, an estimated 347 [@Teh:2015gd] to 633 (Chapter 2) fish stocks have distributions that cross national borders, some of them jointly managed by two or more countries. These species are responsible for more than 70% of these countries total fish catches (Chapter 2). Under Article 63, UNCLOS incentives actions to cooperate on the management of shared stocks [@UnitedNations:1986tl] as management success often depends on effective cooperation between parties [@Munro:2002uf; @Sumaila:2013hv]. Since the definition of shared stocks, game theory has been one of the most common approaches used to analyze the management of these type of stocks. However, shared stocks' management can be convoluted due to the participation of several fishing “players”, different countries and sometimes jurisdictions within a country, the migration patterns of the stock, and their abundance fluctuation within space and time [@Munro:2002uf, Engler and Saunders *in press*]. In addition, international treaties might not be prepared to address the effects that climate change will bring to shared fish stocks (Engler and Saunders *in press*, Koubrak and VanderZwaag, *in press*).

<!-- ##### Climate Change and Transboundary Fisheries -->

The ocean is getting warmer [@Rheim:2013fv], less oxygenated [@Schmidtko:2017cl], and increasing in acidity [@Ross:2011kd; @IPCC:2019tn]. To cope with these changes in ocean biophysical properties, marine species, including shared fish stocks, have been shifting their distribution towards the poles and/or deeper waters [@Poloczanska:2016kk]. As climate change reshapes the ocean's environment worldwide [@Gattuso:2015jz], shared fisheries' governance is threatened as new migration patterns may arise [@Miller:2013iv; @Pinsky:2018cb], historic distribution and abundances might shift [@Cheung:2010dt], and species basic natural traits may modify [@Pauly:2017hpa]. Catches of shared stocks like tunas, have significantly increased in some regions such as the subtropical Atlantic and western Pacific Oceans and are projected to continue [@MonllorHurtado:2017cm; @ErauskinExtramiana:2019fw]. Multiple shared species in North America have been observed to shift in distribution following changes in optimal conditions such as sockeye salmon (*Oncorhynchus nerka*) [@McDaniels:2010vn; @Miller:2013iv; @Song:2017va], Atlantic cod [@Pershing:2015gq], and flounders [@Pinsky:2012kq]. Moreover, these shifts are projected to continue towards the end of the $21^{st}$ century [@Cheung:2018hz]. As a result, some countries or management jurisdictions may see more shared fisheries and their catches shifting into their waters while others will stand to lose [@Oremus:2020id; @Pinsky:2018cb, Chapter 3]. Nevertheless, management rules for shared stocks (e.g., quota or spatial delimitation) are often determined based on current and/or historic knowledge of the stock's distribution and do not consider future shifts in distributions [@FredstonHermann:2018kp].

@@

It is expected that climate induced shifts in stocks’ distribution will affect the rules in place that keep international treaties alive. Therefore, the main objective of this article is to assess the level of exposure that bi-lateral transboundary fisheries treaties between Canada and the US have to climate change through shifts in stock distributions. Specifically, I rely on a species distribution model and scenario planning to project the changes in the distribution of selected fish stocks jointly managed by Canada and the United States focusing on two case studies (the International Pacific Halibut Commission and a fisheries arrangement for the Gulf of Maine) Finally, I explore similar situations around the world and identify opportunities to improve the adaptability of transboundary stocks management to climate change in North America. Despite an overall expectation of species following a poleward shift, important geographic constraints (e.g., Gulf of Alaska representing a latitudinal block) [@Kleisner:2016je], geo-political features (e.g., the localization of Alaska in reference to Canada and the contiguous states), and management rules (e.g., quota allocations, spatial management rules) may play an important role in the redistribution of benefits. Understanding these stocks shifts will shed a light on future conditions and inform decision-makers on the paths to follow under a changing climate.

@@

The current study used the International Pacific Halibut Commission (IPHC) and the Gulf of Maine arrangement (hereafter referred as GoMa) as case studies to discuss the implications that climate change could have in the management of transboundary stocks. For the IPHC, I used the most updated spatial regulatory data along its 12 regulatory areas [@IPHCSecreatriat:2017uz; @IPHC:2019wt]. For this specific case, I considered Alaska as a separate entity, the US contiguous states as a second one (Washington, Oregon and California), and lastly British Columbia (Canada). For the GoMa I used the Northwest Atlantic Fisheries Organization's (NAFO)[^NAFO] divisions 5Y, 5Ze, and 4X within latitudes 46.2°N and 41.5°S, and longitudes -72°W and -64°E (Figure \@ref(fig:ocFigure1)). Is worth mention that, while NAFO's divisional zones were used in this study for domestic management, NAFO does not manage fisheries within the EEZs of Canada and the United States. Fisheries data for the projections was gathered from the Sea Around Us from 1951 to 2014 [@Zeller2016].

@@

The IPHC was established in 1923 by Canada and the United States to oversee the management of Pacific halibut (*Hippoglossus stenolepis*) [@IPHC:2014tz]. There are 12 regulatory areas from which 3AB holds 51.2% of the stock, followed by regions 2ABC and 4ACDE with 23.1% and 20.4%, respectively, and lastly region 4B with only 5.2% of the stock distribution [@IPHC:2018vz]. In terms of management, the IPHC implements a total allowable catch (TAC) based on a yearly sampling of the Convention area in addition to a series of regulations to control fishing effort [@IPHC:2018vz]. The TAC is divided between recreational, subsistence and commercial fishery, with a portion set aside for bycatch of other fisheries [@IPHC:2019wt]. The commercial fishing season starts in March ending around November with restrictions allowing only set line gear with J-type hooks targeting individuals over 81.3 cm of total length [@IPHC:2019wt].

@@

The results of the present study suggest that climate change will alter the MCP of jointly managed transboundary fish stocks in North America consequently altering Canada’s and the US’s stock’s stock-share ratio, regardless of the climate change scenario. These results are aligned with regional [@Morley:2018fn] projections suggesting that climate change will push marine species towards the poles and deeper water [@Pinsky:2013ab] in search of their ecological niche [@Poloczanska:2016kk]. Moreover, IPHC data[^IPHC] suggest that some of these shifts are already happening. For example, since 2010, the distribution proportion of Pacific halibut has increased from 9% to 11% in region 2B, from 7.5% to 13% in region 2C, and from 12.3% to 13.5% in region 4CDE. On the other hand, regions 3A and 3B have seen the largest decreases in the IPHC regulatory areas since 2010, from 35.3% to 30.6% and 20.6% to 15.9%, respectively. Similarly, in the Gulf of Maine, the projected stock-share gain of yellowtail flounder and haddock by the US (Figure \@ref(fig:ocFigure5)) follows a historical trend where in 2019, Canada's stock-share decreased from 35% to 32% and 60% to 40% relative to 2010, respectively [@Trinko:2019ou].

@@

Shifts in stocks distribution due to climate change have the potential of creating local extinction of economically important stocks while enhancing fisheries in areas where they were not present before. In this chapter, I have explored the potential impacts of climate change in the joint management of selected transboundary stocks managed by Canada and the US. I found that, transboundary stocks are likely to shift in the upcoming years changing the proportion of the catch of jointly managed fisheries of Canada and the US. Lessons from other countries can provide solutions to such challenges. More specific, side payments, dynamic management, and interchangeable quotas were identified as potential solutions for North American region. While not directly addressed in this study, socio-economic impacts of shifting transboundary stocks could add an extra layer of complexity to the problem. Addressing shifts in stocks distribution sooner rather than latter could avert the so called “fish wars”, improve sustainability of jointly managed stocks, and secure the livelihood of thousands of families that depend on stocks that move freely between national jurisdictions. Finally, preparing for an uncertain future is key to achieve sustainable fisheries.

>>>>

In 1982, the United Nations Law of the Sea Convention (UNCLOS) formalized the concept of exclusive economic zones (EEZs) creating the concept that we know today as shared stocks [@UnitedNations:1986tl], i.e., stocks that migrate between countries EEZs (known as transboundary stocks) or between EEZs and the high seas (also called straddling stocks) [@Song:2017iua]. Today, an estimated 347 [@Teh:2015gd] to 633 (Chapter 2) fish stocks have distributions that cross national borders, some of them jointly managed by two or more countries. These species are responsible for more than 70% of these countries total fish catches (Chapter 2). Under Article 63, UNCLOS incentives actions to cooperate on the management of shared stocks [@UnitedNations:1986tl] as management success often depends on effective cooperation between parties [@Munro:2002uf; @Sumaila:2013hv]. Since the definition of shared stocks, game theory has been one of the most common approaches used to analyze the management of these type of stocks. However, shared stocks' management can be convoluted due to the participation of several fishing “players”, different countries and sometimes jurisdictions within a country, the migration patterns of the stock, and their abundance fluctuation within space and time [@Munro:2002uf; @Engler2020]. In addition, international treaties might not be prepared to address the effects that climate change will bring to shared fish stocks [@Engler2020; @Kubrack2020]
<!-- ##### Climate Change and Transboundary Fisheries -->

The ocean is getting warmer [@Rheim:2013fv], less oxygenated [@Schmidtko:2017cl], and increasing in acidity [@Ross:2011kd; @IPCC:2019tn]. To cope with these changes in ocean biophysical properties, marine species, including shared fish stocks, have been shifting their distribution towards the poles and/or deeper waters [@Poloczanska:2016kk]. As climate change reshapes the ocean's environment worldwide [@Gattuso:2015jz], shared fisheries' governance is threatened as new migration patterns may arise [@Miller:2013iv; @Pinsky:2018cb], historic distribution and abundances might shift [@Cheung:2010dt], and species basic natural traits may modify [@Pauly:2017hpa]. Catches of shared stocks like tunas, have significantly increased in some regions such as the subtropical Atlantic and western Pacific Oceans and are projected to continue [@MonllorHurtado:2017cm; @ErauskinExtramiana:2019fw]. Multiple shared species in North America have been observed to shift in distribution following, in part, changes in optimal conditions such as sockeye salmon (*Oncorhynchus nerka*) [@McDaniels:2010vn; @Miller:2013iv; @Song:2017va], Atlantic cod [@Pershing:2015gq], and flounders [@Pinsky:2012kq]. Moreover, these shifts are projected to continue towards the end of the $21^{st}$ century [@Cheung:2018hz]. As a result, some countries or management jurisdictions may see more shared fisheries and their catches shifting into their waters while others will stand to lose [@Oremus:2020id; @Pinsky:2018cb, Chapter 3]. Nevertheless, management rules for shared stocks (e.g., quota or spatial delimitation) are often determined based on current and/or historic knowledge of the stock's distribution and do not consider future shifts in distributions [@FredstonHermann:2018kp].

@@

It is expected that climate induced shifts in stocks’ distribution will affect the rules in place that keep international treaties alive. Therefore, the main objective of this article is to assess the level of exposure that bi-lateral transboundary fisheries treaties between Canada and the US have to climate change through shifts in stock distributions. Specifically, I rely on a species distribution model and scenario planning to project the changes in the distribution of selected fish stocks jointly managed by Canada and the US focusing on two case studies (the International Pacific Halibut Commission and a fisheries arrangement for the Gulf of Maine) Finally, I explore similar situations around the world and identify opportunities to improve the adaptability of transboundary stocks management to climate change in North America. Despite an overall expectation of species following a poleward shift, important geographic constraints (e.g., Gulf of Alaska representing a latitudinal block) [@Kleisner:2016je], geo-political features (e.g., the localization of Alaska in reference to Canada and the contiguous states), and management rules (e.g., quota allocations, spatial management rules) may play an important role in the redistribution of benefits. Understanding these stocks shifts will shed a light on future conditions and inform decision-makers on the paths to follow under a changing climate.

@@

In this chapter, I used the International Pacific Halibut Commission (IPHC) and the Gulf of Maine arrangement (hereafter referred as GoMa) as case studies to discuss the implications that climate change could have in the management of transboundary stocks. For the IPHC, I used the most updated spatial regulatory data along its 12 regulatory areas [@IPHCSecreatriat:2017uz; @IPHC:2019wt]. For this specific case, I considered Alaska as a separate entity, the US contiguous states as a second one (Washington, Oregon and California), and lastly British Columbia (Canada). For the GoMa, I used the Northwest Atlantic Fisheries Organization's (NAFO)[^NAFO] divisions 5Y, 5Ze, and 4X within latitudes 46.2°N and 41.5°S, and longitudes -72°W and -64°E (Figure \@ref(fig:ocFigure1)). It is worth mention that, while NAFO's divisional zones were used in this study for domestic management, NAFO does not manage fisheries within the EEZs of Canada and the United States. Fisheries data for the projections was gathered from the Sea Around Us from 1951 to 2014 [@Zeller2016].

@@

The IPHC was established in 1923 by Canada and the US to oversee the management of Pacific halibut (*Hippoglossus stenolepis*) [@IPHC:2014tz]. There are 12 regulatory areas from which 3AB holds 51.2% of the stock, followed by regions 2ABC and 4ACDE with 23.1% and 20.4%, respectively, and lastly region 4B with only 5.2% of the stock distribution [@IPHC:2018vz]. In terms of management, the IPHC implements a total allowable catch (TAC) based on a yearly sampling of the Convention area in addition to a series of regulations to control fishing effort [@IPHC:2018vz]. The TAC is divided between recreational, subsistence and commercial fishery, with a portion set aside for bycatch of other fisheries [@IPHC:2019wt]. The commercial fishing season starts in March ending around November with restrictions allowing only set line gear with J-type hooks targeting individuals over 81.3 cm of total length [@IPHC:2019wt].

@@

The results of the present chapter suggest that climate change will alter the MCP of jointly managed transboundary fish stocks in North America consequently altering Canada’s and the US’s stock’s stock-share ratio, regardless of the climate change scenario. These results are aligned with regional [@Morley:2018fn] projections suggesting that climate change will push marine species towards the poles and deeper water [@Pinsky:2013ab] in search of their ecological niche [@Poloczanska:2016kk]. Moreover, IPHC data[^IPHC] suggest that some of these shifts are already happening. For example, since 2010, the distribution proportion of Pacific halibut has increased from 9% to 11% in region 2B, from 7.5% to 13% in region 2C, and from 12.3% to 13.5% in region 4CDE. On the other hand, regions 3A and 3B have seen the largest decreases in the IPHC regulatory areas since 2010, from 35.3% to 30.6% and 20.6% to 15.9%, respectively. Similarly, in the Gulf of Maine, the projected stock-share gain of yellowtail flounder and haddock by the US (Figure \@ref(fig:ocFigure5)) follows a historical trend where in 2019, Canada's stock-share decreased from 35% to 32% and 60% to 40% relative to 2010, respectively [@Trinko:2019ou].

@@

Shifts in stocks distribution due to climate change have the potential of creating local extinction of economically important stocks while enhancing fisheries in areas where they were not present before. In this chapter, I have explored the potential impacts of climate change in the joint management of selected transboundary stocks managed by Canada and the US. I found that, transboundary stocks are likely to shift in the upcoming years changing the proportion of the catch of jointly managed fisheries of Canada and the US. Lessons from other countries can provide solutions to such challenges. More specific, side payments, dynamic management, and interchangeable quotas were identified as potential solutions for North American region. While not directly addressed in this chapter, socio-economic impacts of shifting transboundary stocks could add an extra layer of complexity to the problem. Addressing shifts in stocks distribution sooner rather than latter could avert the so called “fish wars”, improve sustainability of jointly managed stocks, and secure the livelihood of thousands of families that depend on stocks that move freely between national jurisdictions. Finally, preparing for an uncertain future is key to achieve sustainable fisheries.
